Efficient implementation of kernel filtering

### 2007-06-23 article/Article.tex
Fixed border issues, as far as possible.
darcs-hash:20070622074102-fafb9-babf2e36ef40df789d6caa002bb4b6dd4fac545f.gz
started performance section in article
darcs-hash:20070623130246-fafb9-d0d75cfd007716fb0ffc65bd869ce393939d23d4.gz

\subsection{Arbitary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps.

@@

\section{Rank filters}
Efficient implemetations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitary kernel shapes and pixel types as well as any
choice of rank.

@@

\section{Mean and variance filters}
A simplified version of the moving histogram approach can also be used
to efficiently compute the mean or variance of an arbitrarily shaped
kernel. The histogram used for morphology and rank filters is a data
structure that, once established for one pixel, can be easily updated
to represent the kernel for a neighboring pixel. The data structure
required to perform the equivalent function for mean and variance
operations is much simpler -- all that is necessary is the sum of
pixel values under the kernel (or sum of squared values for variance
calculations), and the kernel size. The sum can be updated by adding
the new values and subtracting the old and the output computed by
dividing by the kernel size.

@@

\section{ITK implementation}

>>>>

\subsection{Arbitary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps. This is an extension to the
original paper.

@@

\section{Rank filters}
\label{sect:rank}
Efficient implemetations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitary kernel shapes and pixel types as well as any
choice of rank.

@@

\section{Mean and variance filters}
\label{sect:meanVar}
A simplified version of the moving histogram approach can also be used
to efficiently compute the mean or variance of an arbitrarily shaped
kernel. The histogram used for morphology and rank filters is a data
structure that, once established for one pixel, can be easily updated
to represent the kernel for a neighboring pixel. The data structure
required to perform the equivalent function for mean and variance
operations is much simpler -- all that is necessary is the sum of
pixel values under the kernel (or sum of squared values for variance
calculations), and the kernel size. The sum can be updated by adding
the new values and subtracting the old and the output computed by
dividing by the kernel size.

@@

\section{ITK implementation}
The filters discusses in this article are contained in 3 packages --
{\em consolidatedMorphology}, {\em fastRankMean} and {\em
boxConvolution}. The most recent versions of these packages can be
obtained from {\bf link to Gaetan's darcs repo}. All of the packages
include multithreaded filters and include many examples.

\subsection{Morphology filters -- consolidatedMorphology}
An earlier version of the {\em consolidatedMorphology} package has
been submitted to the InsightJournal. However many improvements and
bug fixes have been included since then.

The morphology filters utilize a new class -- {\em
FlatStructuringElement} to describe, arbitary and decomposable
structuring elements. It provides the following methods to create
structuring elements with different characteristics:
\begin{itemize}
  \item \verb$Ball()$ generate a ball structuring element. It takes the 
	radius of the structuring element as parameter.
  \item \verb$Box()$ produces a box structuring element. It takes the radius 
	of the structuring element as parameter.
  \item \verb$Poly()$ produces an approximation to a circle or sphere using 
	line structuring elements, taking the desired radius and number of 
	lines as parameters. In the 3D case an approximation of a sphere will 
	be attempted. Caution is advised because this is not well tested yet.
  \item \verb$FromImage()$ produces a structuring element from an image. The image is passed
as parameter to the method. A optional parameter can also be passed: the pixel value
to be considered as the foreground value in the image. This value defaults to
\verb$NumericTraits< PixelType >::max()$. This lets the user visualize the structuring element.
\end{itemize}

The algorithm used by a filter is set by the {\em SetAlgorithm}
method, with options of {\em BASIC}, {\em HISTO}, {\em VHGW} and {\em
ANCHOR} for the traditional ITK implementation, the moving histogram
algorithm, the van Herk, Gil, Werman algorithm and the anchor
algorithm respectively. The VHGW and ANCHOR algorithms can only be
used for Box and Poly structuring elements. VHGW is more thoroughly
tested.

\subsection{Rank and mean filters -- fastRankMean}
The {\em fastRankMean} packge using the sliding window method to
implement rank and mean filters for arbitary kernel shapes. It also
uses line versions of these kernels to provide a ``separable rank''
filter -- an fast approximation to a real rank filter as discussed in
Section \ref{sect:rank}. There are also a number of filters that
accept a mask input and return the rank or mean value of the
intersection of kernel and mask. These are experimental and haven't
been thoroughly tested.
\section{Performance}
\subsection{Box Convolution}
The relative times for optimized and standard convolution are shown in
Table \ref{tab:perfBoxConv}, and demonstrate significantly improved
performance sor kernel radii greater than 3. The execution times of
box convolution should be relatively constant with kernel size, but an
obvious increase is clear. This is attributable to the process of
dealing with the boundary conditions, which utilizes shaped
neighborhood filters. This aspect of the implementation needs further work

\begin{table}[htbp]
\centering
\begin{tabular}{ccc}
\hline
Size  &  Direct &  Box \\
\hline
\hline
1   &    0.0109048   &    0.0233631 \\
2   &    0.022008    &    0.0245971 \\
3   &    0.0434739   &    0.0263161 \\
4   &    0.0780701   &    0.0283061 \\
5   &    0.129174    &    0.0310797 \\
6   &    0.199876    &    0.0344833 \\
7   &    0.289754    &    0.0382629 \\
8   &    0.407095    &    0.043121 \\
9   &    0.552862    &    0.0489335 \\
10  &    0.728667    &    0.0555441 \\
15  &    2.17149     & 0.107776 \\
20  &    4.83458     & 0.199508 \\
25  &    9.12699     & 0.372637 \\
\hline

\end{tabular}
\caption{Execution times for direct and box convolution implementations applied to the $256 \times 256$ cthead image.\label{tab:perfBoxConv}}
\end{table}
